Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-02-08 text/paper2/ms.tex
lots of writin'

[TODO: A historical introduction to dark matter, potential measures?]

Major open questions in near-field cosmology: (1) is the Milky Way potential triaxial and if so, how much? (2) How does the potential vary with radius? (3) Is there graininess to the potential? e.g., is there substructure?

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm seem to produce triaxial dark matter halos \citep{dubinski91, jing02, kuhlen07, veraciro11}. Halos tend to be more strongly triaxial in dark-matter-only simulations, but still present with the inclusion of baryons. Baryonic processes tend to modify the dark matter structure in the inner galaxy, leading possibly to radially-dependent axis ratios and halo orientation. [statements about integrability]. [if want to connect lambda-CDM to the milky way, we need to face non-integrable potentials]

[something about measuring halo shape with lensing].

[why use tidal streams?] \cite{lm10} used N-body models to ... triaxial.

Another prediction of hierarchical structure formation in $\Lambda$CDM is the presence of substructure.

- Broader comparison to other methods\\
\indent\indent- 'generative methods' more generally (ie N-body, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- statistical methods\\
\indent\indent- Sanders \& Binney 2013\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

>>>>

Cosmological simulations of galaxy formation in the $\Lambda$CDM paradigm predict dark matter halos that (1) are permeated with substructure on many scales, (2) are triaxial in shape, and (3) have density profiles and shapes that vary with radius \citep{dubinski91, jing02, kuhlen07, veraciro11}. Dark-matter-only simulations seem to produce strongly triaxial halos, but including baryons tends to soften the triaxiality in the inner galaxy [through a combination of adiabatic contraction and energetic feedback \citep[see recent discussion by][]{pontzen12} TODO: modify previous sentence bc in PW\&J 2013]. These [...] may lead to radially-dependent axis ratios, halo orientation, and density profiles.

The bulk of the baryonic matter in galaxies spans roughly 5-10\% of the extent of the host dark matter halo. Hence, the major components of a galaxy are sensitive to the inner portion of its host halo's mass distribution. For example, the rotation curves of disk galaxies are particularly sensitive probes of the inner mass since matter in disks can be assumed to move on near circular orbits. This extra knowledge about the orbital distribution of the baryonic matter within a disk allows a more precise tracing of the potential than is feasible for objects such as elliptical galaxies where orbits are assumed to be random. [Potential is most likely smoothed out (energetic feedback?), closely integrable allowing beautiful dynamical ...]

Conversely, measuring the dark matter distribution at large radii is difficult as the density of visible tracers becomes small. Around external galaxies, the extended mass distribution has been explored using a variety of approaches \citep[see][for a a complete and detailed review]{courteau12}. For example, the kinematics of tracer populations such as globular clusters or planetary nebulae can be used to derive mass estimates under the assumptions that these satellite systems are relaxed and well-mixed \citep[early investigations include][]{mendez01,cote03}. These estimates are necessarily limited by the number of objects bright enough for spectroscopic measurements and thus the largest samples to date have roughly several hundred members probing out to several 10's of kpc. Simple, parameterized models of both the mass and orbit distribution have been simultaneously constrained using such data \citep[e.g.][]{napolitano11,deason12a}. 
%\citep{romanowsky03,douglas07,romanowsky09,napolitano11,lee11}
Alternatively, the statistical properties of gravitationally lensed background sources around a galaxy can be used to constrain the \emph{projected} shape, orientation, and radial profile of mass \citep[as done by the Lens Structure and Dynamics Survey described in][]{koopmans02}. Of course, lensing reconstructions can only be performed for galaxies which closely intersect our line of sight to background sources, but the advent of large photometric catalogues has allowed automatic searches for such chance alignments and significant increases in the number of objects studied in this way \citep[e.g. the Sloan Lens ACS Survey, see][]{bolton06}.

Within the Milky Way we have a unique vantage point that allows us to measure the three-dimensional structure of our own dark matter halo. One approach to ... [explain ``random'' methods: kinematic tracers]. Another approach is to exploit the non-random nature of the stellar populations in the halo. A consequence of hierarchical structure formation [...] and such structures have been observed [e.g., ...cite belokurov?]. Tidal streams are dynamically cold systems --- debris typically have small distributions of energy and angular momenta --- and thus require orders of magnitude fewer tracers than a random sample to get constraints of comparable accuracy. For example, in the simplest case we might {\it assume} that debris stars are actually still on the same orbit as their progenitor system \citep[a \emph{wrong} assumption, see e.g.][]{eyre}. Then we can imagine measuring the full-space velocities ${\bf v}$ at different points ${\bf x}$ in the structure (e.g. along a stream), which would give us a direct measure of differences in a potential, $\Phi$.

Since the initial discovery of tidal debris from Sgr, myriad other streams and debris systems have been discovered in large photometric surveys probing Galactocentric distances from $\sim$15-100 kpc. This has motivated numerous efforts for developing methods to constrain properties of the Galactic potential with cold debris structures in the Milky Way halo.

- Broader comparison to other methods\\
\cite[][hereafter LM10]{lm10} ran N-body models of the disruption of the Sagittarius (Sgr) dwarf spheroidal galaxy and used these simulations to constrain the shape of the Milky Way potential out to $\sim$70 kpc [check number], finding that the halo ... (though, see ... for a discussion of unexplained features in the Sgr debris).
\indent\indent- 'generative methods' more generally (ie N-body, plus Varghese, plus Kuepper plus new possible action-angle models....\\
\indent\indent- statistical methods\\
\indent\indent- Sanders \& Binney 2013\\
- Then a short note on worries\\
\indent\indent- for generative models\\
\indent\indent- for any methods based on action/angle \\
\indent\indent- GD-1 was fit as if it highlighted an orbit.  That is known to be wrong.\\
\indent\indent- Densities are hard to match: need a background model, way to go from luminosity to density, needs assumptions about the mass function and structure of the satellite\\
\indent\indent- Triaxiality / radially-dependent shapes and substructure present problems for integral-based methods\\
- Then how rewinder gets round these????\\
\indent\indent- Reference Deg \& Widrow to say they illustrate the idea of the power of multiple streams to back up the bit where you say that's the long-term aim\\

[Another prediction of hierarchical structure formation in $\Lambda$CDM is the presence of substructure. ]
